---
title: "How severe weather events affect population health and the economy"
author: Hedley Stirrat
output: html_document
---
## Synopsis
This report examines the effects of severe weather events on population health
and the economy. We use the Storm Data dataset of the National Oceanic and Atmospheric Administration, comprising
observations from 1950 to 2011 in the United States.
Two primary questions are addressed in this report:

1. Across the United States, which types of events are most harmful with respect
to population health?

2. Across the Unites States, which types of events have the greatest economic
consequences?

It was found that tornados have caused the greatest number of fatalities and
injuries in the Unites States between 1950 and 2011. On the other hand, since
the mid-1990s, floods have caused the most damage to property and crops,
followed by hurricanes/typhoons and tornados.

## Data processing
- The data was accessed directly from
`https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2`
on 18 April, 2020. 

### Load packages and read in data
- After importing the data, we observe 902297 observations of 37
variables
- Return brief summaries of the `FATALITIES` and `INJURIES` variables
- `PROPDMGEXP` and `CROPDMGEXP` variables appear to be modifiers that determine
the multiplication to apply to `PROPDMG` and `CROPDMG`, respectively, to obtain
the real numeric value. Check what type of values are present in these variables

```{r}
# load packages
library(dplyr) # version 0.8.5
library(ggplot2) # version 3.3.0
library(lubridate) # version 1.7.4

# read in data
df <- read.csv('repdata_data_StormData.csv.bz2')

# check dimensions
dim(df)

# summarise fatalities and injuries
summary(df$FATALITIES)
summary(df$INJURIES)

# check for `PROPDMGEXP` and `CROPDMGEXP` values
unique(df$PROPDMGEXP)
unique(df$CROPDMGEXP)

# see how many unique `EVTYPE` values there are
length(unique(df$EVTYPE))
```

### Subset and transform data
- Select only the relevant variables for this analysis
- Convert `BGN_DATE` to a Date-type variable, add associated `Year` variable
- Convert `PROPDMG` and `PROPDMGEXP` into a single PropertyDamage
numeric variable, and apply an analogous transformation to `CROPDMG` and
`CROPDMGEXP`
- Combine `PropertyDamage` and `CropDamage` into a single `MonetaryLoss` variable
- Convert all `EVTYPE` to upper case, so that they can be better grouped

```{r}
df <- df %>%
    select(
        BGN_DATE,
        EVTYPE,
        FATALITIES,
        INJURIES,
        PROPDMG,
        PROPDMGEXP,
        CROPDMG,
        CROPDMGEXP) %>%
    mutate(
        BGN_DATE = as.Date(BGN_DATE, format = '%m/%d/%Y %H:%M:%S'),
        Year = year(BGN_DATE),
        PropertyDamage = case_when(
            PROPDMGEXP == 'H' ~ PROPDMG * 100,
            PROPDMGEXP == 'h' ~ PROPDMG * 100,
            PROPDMGEXP == 'K' ~ PROPDMG * 1000,
            PROPDMGEXP == 'k' ~ PROPDMG * 1000,
            PROPDMGEXP == 'M' ~ PROPDMG * 1000000,
            PROPDMGEXP == 'm' ~ PROPDMG * 1000000,
            PROPDMGEXP == 'B' ~ PROPDMG * 1000000000,
            PROPDMGEXP == 'b' ~ PROPDMG * 1000000000,
            PROPDMGEXP == '' ~ PROPDMG,
            TRUE ~ PROPDMG),
        CropDamage = case_when(
            CROPDMGEXP == 'H' ~ CROPDMG * 100,
            CROPDMGEXP == 'h' ~ CROPDMG * 100,
            CROPDMGEXP == 'K' ~ CROPDMG * 1000,
            CROPDMGEXP == 'k' ~ CROPDMG * 1000,
            CROPDMGEXP == 'M' ~ CROPDMG * 1000000,
            CROPDMGEXP == 'm' ~ CROPDMG * 1000000,
            CROPDMGEXP == 'B' ~ CROPDMG * 1000000000,
            CROPDMGEXP == 'b' ~ CROPDMG * 1000000000,
            TRUE ~ CROPDMG),
        MonetaryLoss = PropertyDamage + CropDamage,
        EVTYPE = toupper(EVTYPE))

# converting to upper case has condensed the unique number of EVTYPE values
length(unique(df$EVTYPE))
```

### Strategy for ranking event types
- The questions of which event types are associated with higher number of
fatalities and injuries, and worse economic consequences, could be addressed in
a couple of different ways. We could look at the
median or mean number of injuries and fatalities (or economic loss) per
event type to find the events with the highest average consequences;
alternatively, we could take the approach of finding the event types with the
highest total number of injuries and fatalities (or economic loss).
- In this
report, we have chosen the second option. This is because there are around 900
unique event types in the dataset, but the majority of these event types have
only one, two or a few observations. Consider the following summary:

```{r}
df %>%
    group_by(EVTYPE) %>%
    summarise(EventTypeInstances = n()) %>%
    ungroup %>%
    group_by(EventTypeInstances) %>%
    summarise(Frequency = n())
```

- There are 456 event types that occur only once in the data. Without filtering
these out, mean and median calculations for event types could be misleading.
Therefore, this analysis instead looks at summed instances of fatality and
injury (and property/crop damage) to determine those events with the greatest consequences.


## Results

### Harm to population health: fatalities and injuries
- To determine which types of storm events are associated with the highest
number of injuries and fatalities, we group the data by event type, sum the
number of injuries and fatalities per event type, take the ten highest
categories, and plot these.
- Figure 1 compares the ten event types with the highest summed fatalities in the US between 1950-2011, while Figure 2 compares the ten event types with the highest summed injuries. The 'Tornado' event type has caused the greatest total number of fatalities,
and by a very large margin, the greatest number of injuries.

```{r}
df %>%
    group_by(EVTYPE) %>%
    summarise(total_fatalities = sum(FATALITIES)) %>%
    arrange(desc(total_fatalities)) %>%
    head(10) %>%
    arrange(total_fatalities) %>%
    mutate(Position = factor(EVTYPE, EVTYPE)) %>%
    ggplot(aes(x = Position, y = total_fatalities)) +
        geom_bar(stat='identity') +
        coord_flip() +
        ggtitle('Figure 1. Total fatalities per event type') +
        labs(x = 'Event type', y = 'Total fatalities') + 
        theme(plot.title = element_text(hjust = 0.5))

df %>%
    group_by(EVTYPE) %>%
    summarise(total_injuries = sum(INJURIES)) %>%
    arrange(desc(total_injuries)) %>%
    head(10) %>%
    arrange(total_injuries) %>%
    mutate(Position = factor(EVTYPE, EVTYPE)) %>%
    ggplot(aes(x = Position, y = total_injuries)) +
        geom_bar(stat='identity') +
        coord_flip() +
        ggtitle('Figure 2. Total injuries per event type') +
        labs(x = 'Event type', y = 'Total injuries') + 
        theme(plot.title = element_text(hjust = 0.5))
```

### Economic consequences: property and crop damage
- When looking at financial data spanning several decades, it is important to
take inflation into account. In this case, we have data spanning over six
decades; a given dollar amount of crop or property damage in 1950 is
significantly different in real terms to the same dollar amount in 2011. 
- Instead of adjusting for inflation, it might be easier to simply remove a
large portion of historical observations. Indeed, the number of observations per
year remained relatively low up until the mid-1990s; by only using data from
this period onward, we are comparing (relatively) like-for-like monetary losses.
- The sum will be used instead of the mean or median, for reasons discussed
above.
- Figure 3 compares the ten event types with the highest summed crop and property damage consequences, in the US between 1996-2011. Floods have caused the most economic damage in the period investigated,
followed by hurricanes/typhoons and storm surges.

```{r}
df %>% 
    filter(Year > 1995) %>% 
    group_by(EVTYPE) %>%
    summarise(MonetaryLoss = sum(MonetaryLoss)) %>%
    arrange(desc(MonetaryLoss)) %>%
    head(10) %>%
    arrange(MonetaryLoss) %>%
    mutate(Position = factor(EVTYPE, EVTYPE)) %>%
    ggplot(aes(x = Position, y = MonetaryLoss / 1000000)) +
        geom_bar(stat='identity') +
        coord_flip() +
        ggtitle('Figure 3. Total property and crop damage per event type') +
        labs(x = 'Event type', y = 'Total damage ($M USD)') + 
        theme(plot.title = element_text(hjust = 0.5)) +
        scale_y_continuous(labels = scales::comma)
```

