# Quick look at interaction terms in multivariate regression models

This example is motivated by one of the examples in the Johns Hopkins
University's Data Science Specialization (Regression Models course).

## Research question
With the mtcars dataset, fit a model that predicts mpg using the interaction
between the number of cylinders (as factor) and weight.

## Fit model
```{r}
fit <- with(mtcars, lm(mpg ~ wt * factor(cyl)))
```

## Plot
```{r}
# plot data with three lines, each isolating a single cyl

ggplot(mtcars, aes(x=wt,y=mpg)) + geom_point(aes(color=factor(cyl))) +
    # cyl = 4
    geom_abline(intercept = coef(fit)[1], slope=coef(fit)[2]) +
    # cyl = 6
    geom_abline(intercept = coef(fit)[1] + coef(fit)[3], slope = coef(fit)[2] + coef(fit)[5]) +
    # cyl = 8
    geom_abline(intercept = coef(fit)[1] + coef(fit)[4], slope = coef(fit)[2] + coef(fit)[6])
```

## Compare with simpler model that doesn't include interaction
```{r}
fit2 <- with(mtcars, lm(mpg ~ wt + factor(cyl)))

anova(fit,fit2)
```

P-value > 0.05, conclude the interaction term doesn't add anything.
