---
title: "Statistical Inference Course Project Part 2: Tooth Growth Analysis"
author: "Hedley Stirrat"
date: "17 May 2020"
output: pdf_document
---

## Overview
The ToothGrowth dataset in the R datasets package contains the results of an
  experiment that explored the effect of vitamin C on tooth growth in guinea
  pigs. Each guinea pig received one of three dose levels of vitamin C by one of two
  delivery methods.

The ToothGrowth dataframe consists of 60 observations of 3 variables. The variables are
  `len` (tooth length), `supp` (supplement type, VC or OJ for ascorbic acid and
orange juice, respectively), and `dose` (dose in mg/day).

We will conduct an initial exploratory data analysis to generate hypotheses
about the data, and then test those hypotheses in the final section.

## Initial exploratory analysis

Simply plotting dose vs length suggests that higher vitamin C doses may lead to
  longer tooth lengths:

```{r echo=FALSE, fig.height=4}
library(ggplot2)
ggplot(ToothGrowth, aes(x = as.factor(dose), y = len)) +
    geom_point() +
    ggtitle("Effect of vitamin C dose on tooth length") +
    labs(y = "Tooth length", x = "Dose (mg/vitamin C)")
```

It's not clear that the supplement type affects tooth length in an obvious way:

```{r echo=FALSE, fig.height=3}
ggplot(ToothGrowth, aes(x = supp, y = len)) +
    geom_point() +
    ggtitle("Effect of supplement type on tooth length") +
    labs(y = "Tooth length", x = "Supplement type")
```

We can look at these two variables (`supp` and `dose`) and their combined
effect on tooth length. Splitting up the data this way appears to suggest that
at lower dosages (0.5 mg and 1.0 mg), the OJ supplement type leads to better
tooth growth than VC, while at the highest dosage (2.0 mg), supplement type
does not have an effect on tooth growth.

```{r echo=FALSE, fig.height=3}
ggplot(ToothGrowth, aes(x = supp, y = len, colour = supp)) +
    geom_boxplot() +
    facet_wrap(~ dose) +
    ggtitle("Effect of vitamin C dose and supplement type on tooth length") +
    labs(y = "Tooth length", x = "Supplement type", colour = "Type")
```

## Hypothesis testing

### Assumptions
- We have left the default `t.test` parameter of `paired = FALSE` as the data do
  not appear to be paired
- We have left the default `t.test` parameter of `conf.level = 0.95` to test for
  significance at the 95% confidence level
- We have left the default `t.test` parameter of `alternative = "two.sided"` to
  conduct a two-sided test (although one-sided may be more appropriate here)
- We have left the default `t.test` parameter of `var.equal = FALSE` with the
  assumption that the variances between groups are not equal
- We also make the assumption that the variables are independent and identically distributed

### Dosage vs tooth length

The first plot above suggested that higher vitamin C doses may be associated
with longer tooth lengths. Let's test this hypothesis by running t-tests on
unique dosage combinations (0.5 mg and 1.0 mg; 0.5 mg and 2.0 mg; 1.0 mg and
2.0 mg).

```{r echo=FALSE}
dose_vs_len <- list()
dose_vs_len[['0.5-1.0mg']] <- t.test(len ~ dose, data = ToothGrowth[ToothGrowth$dose %in% c(0.5,1.0),])
dose_vs_len[['0.5-2.0mg']] <- t.test(len ~ dose, data = ToothGrowth[ToothGrowth$dose %in% c(0.5,2.0),])
dose_vs_len[['1.0-2.0mg']] <- t.test(len ~ dose, data = ToothGrowth[ToothGrowth$dose %in% c(1.0,2.0),])

generate_summary <- function(result_list) {

    if (length(result_list) == 10) {
        `p-value` <- result_list[['p.value']]
        `lower confidence` <- round(result_list[['conf.int']][[1]], 2)
        `upper confidence` <- round(result_list[['conf.int']][[2]], 2)
    } else {
        `p-value` <- unlist(lapply(result_list, function(x) x[['p.value']]))
        `lower confidence` <- round(unlist(lapply(result_list, function(x) x[['conf.int']][[1]])),2)
        `upper confidence` <- round(unlist(lapply(result_list, function(x) x[['conf.int']][[2]])),2)
    }
    data.frame(`p-value`, `lower confidence`, `upper confidence`)
}

generate_summary(dose_vs_len)
```

In all three comparisons, we have identified that higher doses are associated
with longer tooth lengths, and these differences are significant at the 95%
confidence level. Therefore, we reject the null hypothesis that vitamin C dosage
levels are not associated with a difference in tooth lengths.

### Supplement type vs tooth length

As described above, it appears as though the supplement type may have an effect
on tooth length at low dosages of vitamic C, but not at the higher 2.0 mg
dose. 

Before we test this hypothesis, let's first test whether the different
supplement types are associated with different tooth lengths regardless of
dosage level: 

```{r}
generate_summary(t.test(len ~ supp, data = ToothGrowth))
```

With a p-value of 0.06 and our 95% confidence interval containing zero, we
cannot reject the null hypothesis that the two different supplement types do not
lead to different tooth lengths when we look at data containing all dosage levels.

However, let's see whether we can identify differences in tooth lengths caused
by the different supplement types if we control for dosage levels: 

```{r echo=FALSE}
supp_vs_len <- list()
supp_vs_len[['0.5mg']] <- t.test(len ~ supp, data = ToothGrowth[ToothGrowth$dose %in% c(0.5),])
supp_vs_len[['1.0mg']] <- t.test(len ~ supp, data = ToothGrowth[ToothGrowth$dose %in% c(1.0),])
supp_vs_len[['2.0mg']] <- t.test(len ~ supp, data = ToothGrowth[ToothGrowth$dose %in% c(2.0),])

generate_summary(supp_vs_len)
```

The t-tests support our initial hypotheses; we've shown that at dosages of 0.5
or 1.0 mg, the OJ supplement type is associated with a longer tooth length with
p < 0.01 at the 95% confidence level. Additionally, we found no significant
difference between the supplement types at the 2 mg dosage level, as the 95%
confidence interval included zero with a p-value of 0.96.

## Conclusion

We've shown that at low vitamin C dosages of 0.5 and 1.0 mg, the OC supplement
method was associated with longer tooth lengths, while at 2.0 mg dosage, there
was no significant difference in tooth lengths between the supplement types

We've also shown that, regardless of supplement type, higher vitamin C dosages
are associated with longer tooth lengths.

## Appendix

Code used to set up the analysis:

```{r}
library(ggplot2)
summary(ToothGrowth)
```

Code used in the exploratory data analysis:

```{r fig.show='hide'}
ggplot(ToothGrowth, aes(x = as.factor(dose), y = len)) +
    geom_point() +
    ggtitle("Effect of vitamin C dose on tooth length") +
    labs(y = "Tooth length", x = "Dose (mg/vitamin C)")

ggplot(ToothGrowth, aes(x = supp, y = len)) +
    geom_point() +
    ggtitle("Effect of supplement type on tooth length") +
    labs(y = "Tooth length", x = "Supplement type")


ggplot(ToothGrowth, aes(x = supp, y = len, colour = supp)) +
    geom_boxplot() +
    facet_wrap(~ dose) +
    ggtitle("Effect of vitamin C dose and supplement type on tooth length") +
    labs(y = "Tooth length", x = "Supplement type", colour = "Type")
```

Code used in the hypothesis testing:

```{r}
dose_vs_len <- list()
dose_vs_len[['0.5-1.0']] <- t.test(len ~ dose, data = ToothGrowth[ToothGrowth$dose %in% c(0.5,1.0),])
dose_vs_len[['0.5-2.0']] <- t.test(len ~ dose, data = ToothGrowth[ToothGrowth$dose %in% c(0.5,2.0),])
dose_vs_len[['1.0-2.0']] <- t.test(len ~ dose, data = ToothGrowth[ToothGrowth$dose %in% c(1.0,2.0),])

generate_summary <- function(result_list) {

    if (length(result_list) == 10) {
        `p-value` <- result_list[['p.value']]
        `lower confidence` <- round(result_list[['conf.int']][[1]], 2)
        `upper confidence` <- round(result_list[['conf.int']][[2]], 2)
    } else {
        `p-value` <- unlist(lapply(result_list, function(x) x[['p.value']]))
        `lower confidence` <- round(unlist(lapply(result_list, function(x) x[['conf.int']][[1]])),2)
        `upper confidence` <- round(unlist(lapply(result_list, function(x) x[['conf.int']][[2]])),2)
    }
    data.frame(`p-value`, `lower confidence`, `upper confidence`)
}

generate_summary(dose_vs_len)

generate_summary(t.test(len ~ supp, data = ToothGrowth))

supp_vs_len <- list()
supp_vs_len[['0.5mg']] <- t.test(len ~ supp, data = ToothGrowth[ToothGrowth$dose %in% c(0.5),])
supp_vs_len[['1.0mg']] <- t.test(len ~ supp, data = ToothGrowth[ToothGrowth$dose %in% c(1.0),])
supp_vs_len[['2.0mg']] <- t.test(len ~ supp, data = ToothGrowth[ToothGrowth$dose %in% c(2.0),])

generate_summary(supp_vs_len)
```