---
title: "Statistical Inference Course Project Part 1: Simulation Exercise"
author: "Hedley Stirrat"
date: "17 May 2020"
output: pdf_document
---

## Overview
- In this section, we investigate the exponential distribution in R and compare
  it with the Central Limit Theorem (CLT). The CLT states that the distribution
  of averages of independent, identically distributed rangom variables becomes
  that of a standard normal as the sample size increases.
- To explore the CLT, we will run 1000 simulations of the exponential
  distribution with parameters `n = 40` and `rate = 0.2`. 
- The raw results, containing our collection of 1000 simulations of 40 random
  exponentials, will be stored in the vector `results`. We will also take the
  mean of each simulation and store it in the vector `mean_results`.

```{r echo=FALSE}
set.seed(100)
results <- NULL
mean_results <- NULL
for (i in 1:1000) {
    temp <- rexp(n = 40, rate = 0.2)
    results <- c(results, temp)
    mean_results <- c(mean_results, mean(temp))
}
```

## 1.1. Comparison of sample mean and theoretical mean
- The theoretical mean of the exponential distribution is `1/lambda`
- Here, `lambda = 0.2`, so the theoretical mean is `1 / 0.2 = 5`.
- The mean of our simulated distribution closely aligns with the theoretical
  mean:

```{r echo=FALSE, fig.height=4}
hist(mean_results,
    breaks = 25,
    xlab = "Simulated mean",
    main = "Sample means of 1000 simulations of the exponential distribution")
abline(v = mean(results), lwd=5, col='red')
text(7,70, labels=paste0("Sample mean = ", round(mean(results),5)))
```

## 1.2. Comparison of sample variance and theoretical variance
- The theoretical variance can be calculated as follows: `1 / lambda^2 / n`:
```{r}
1/0.2^2/40
```

- The sample variance can be calculated in R with the `var` function, and
  closely aligns with the theoretical variance:

```{r}
var(mean_results)
```

## 1.3. Demonstration of normal distribution
- To demonstrate that our results---containing 1000 simulated means of 40
  exponentials---are approximately normally distributed, we can plot a simple
  histogram of the simulated means (as we have done above, comparing the sample mean and the
  theoretical mean). However, to better understand this normal distribution, we
  will also plot the raw results (i.e., *not the means*) of our 1000 simulations of
  40 exponentials.
- The top plot is a histogram of the distribution of the results of our 1000
  simulations of 40 exponentials. As expected, the distribution is not Gaussian.
- The bottom plot is a histogram of the 1000 simulated means of our 40
  exponentials. Overlaid on the histogram is a normal distribution line in blue,
  which has been calculated using the theoretical mean and standard deviation.
    - As shown by the plot, the 1000 simulated sample means are approximately
      normally distributed, and centered around the theoretical population mean
      of 5.W
    - Due to the CLT, if we were to increase the number of simulations, our
      results would appear more and more normally distributed.

```{r fig.height=10, echo=FALSE}
x <- seq(
    from = min(mean_results),
    to = max(mean_results),
    length = 40)
par(mfrow = c(2, 1))
hist(results,
    breaks = 25,
    main = "1000 simulations of 40 exponentials",
    xlab = "Distribution of values")
hist(mean_results,
    prob = TRUE,
    breaks = 25,
    main= "Means of 1000 simulations of 40 exponentials",
    xlab = "Distribution of means")
curve(dnorm(x, mean = 5, sd = sqrt((5/sqrt(40))^2)), add=TRUE, col = "blue")
text(x = 5.5, y = 0.59, labels = "Normal distribution in blue")
```

## Appendix

Code used to simulate data:

```{r}
set.seed(100)
results <- NULL
mean_results <- NULL
for (i in 1:1000) {
    temp <- rexp(n = 40, rate = 0.2)
    results <- c(results, temp)
    mean_results <- c(mean_results, mean(temp))
}
```

Code used to create Figure 1:

```{r fig.show='hide'}
hist(mean_results,
    breaks = 25,
    xlab = "Simulated mean",
    main = "Sample means of 1000 simulations of the exponential distribution")
abline(v = mean(results), lwd=5, col='red')
text(7,70, labels=paste0("Sample mean = ", round(mean(results),5)))
```

Code used to create Figure 2:

```{r fig.show='hide'}
x <- seq(
    from = min(mean_results),
    to = max(mean_results),
    length = 40)
par(mfrow = c(2, 1))
hist(results,
    breaks = 25,
    main = "1000 simulations of 40 exponentials",
    xlab = "Distribution of values")
hist(mean_results,
    prob = TRUE,
    breaks = 25,
    main= "Means of 1000 simulations of 40 exponentials",
    xlab = "Distribution of means")
curve(dnorm(x, mean = 5, sd = sqrt((5/sqrt(40))^2)), add=TRUE, col = "blue")
text(x = 5.5, y = 0.59, labels = "Normal distribution in blue")
```