# Literate Statistical Programming
- need a documentation language and a programming language
- knitr supports a variety of documentation languages

## How do I make my work reproducible?
- decide to do it (ideally from the start)
- keep track of things, perhaps with a version control system to track
  snapshots/changes
- use software whose operation can be coded
- don't save output
- save data in nonproprietary formats

## Literate programming: Pros
- text and code all in one place, logical order
- data, results automatically updated to reflect external changes
- code is live--automatic 'regression test' when building a document (e.g., if
  you update the document, you'll know if you've broken something)

## Literate programming: Cons
- text and code all in one place; can make documents difficult to read,
  especially if there is a lot of code
- can substantially slow down processing of documents (although there are tools
  to help)

## What is knitr?
- R package written by Yihui Xie
- supports RMarkdown, LaTeX and HTML as documentation languages
- can export to pdf, HTML
- build right into RStudio

## Requirements
- a recent version of R
- a text editor
- some support packages from CRAN
- some knowledge of Markdown, LaTeX or HTML

## Markdown
- a simplified version of 'markup' languages
- no special editor required
- simple, intuitive formatting elements

## What is knitr good for?
- manuals
- short/medium-length technical documents
- tutorials
- reports (especially if generated periodically)
- data preprocessing documents/summaries

## What is knitr not good for?
- very long research articles
- complex time-consuming computations
- documents that require precise formatting

## In R itself...
```{r}
library(knitr)
knit2html('document.Rmd')
browseURL('document.html')
```

## A few notes
- knitr will fill a new document with filler text; delete it
- code chunks begin with ```{r} and end with ```
- all R code goes in between these markers
- code chunks can have names, which is useful when we start making graphics
```{r firstchunk}
### R code here
```
- by default, code in a code chunk is echoed, which means it'll print the
  results of the computation (if there are results to print)

## Processing of knitr documents
- you write the RMarkdown doc (.Rmd)
- knitr produces a Markdown doc (.md)
- knitr converts the Markdown doc into HTML (by default)
- .Rmd --> .md --> .html
- you should not edit or save the .md or .html documents until you are finished

## Ex
- to not output the code, echo=FALSE
```{r name, echo=FALSE}```

- to not show the results, results="hide"
```{r name, results="hide"}```

- inline text computations
```{r computetime, echo=FALSE}
time <- format(Sys.time(), "%a %b %d %X %Y")
rand = rnorm(1)
```
The current time is `r time`. My favourite random number is `r rand`.

- incorporating graphics
Let's first simulate some data.
```{r simulatedata,echo=TRUE}
x <- rnorm(100); y <- x + rnorm(100, sd=0.5)
```{r}
Here is a scatterplot of the data.
```{r scatterplot,fig.height=4}
par(mar = c(5,4,1,1), las=1)
plot(x,y,main='My simulated data')
```{r}

- making tables with xtable
```{r fitmodel}
library(datasets)
data(airquality)
fit <- lm(Ozone ~ Wind + Temp + Solar.R, data=airquality)
```
Here is a table of regression coefficients.
```{r showtable, results='asis'}
library(xtable)
xt <- xtable(summary(fit))
print(xt, type='html')
```

## Setting global options
- sometimes we want to set options for every code chunk that are different from
  the default options
- e.g., we may want to suppress all code echoing and results output
- we have to write some code to set these global options
```{r setoptions,echo=FALSE}
opts_chunk$set(echo=FALSE, results="hide")
```

## Some common options
- output
  - results: 'asis', 'hide'
  - echo: TRUE, FALSE
- figures
  - fig.height: numeric
  - fig.width: numeric

## Caching computations
- what if one chunk takes a long time to run?
- all chunks have to be recomputed every time you reknit the file
- the cache=TRUE option can be set on a chunk-by-chunk basis to store results of
  computation
- after the first run, results are loaded from cache

## Caching caveats
- if the data or code (or anything external) changes, you need to rerun the
  cached code chunks
- dependencies are not checked explicitly
- chunks with significant side effects may not be cacheable