# Reproducible Research: Course Project 1

## Load libraries and preprocess the data
```{r}
library(dplyr, warn.conflicts=FALSE)
library(tidyr)
library(ggplot2)

df <- read.csv('activity.csv') %>%
    mutate(date = as.Date(date, format="%Y-%m-%d"))
```

## Mean total number of steps per day
### 1. Histogram of total number of steps per day
```{r}
df %>%
    drop_na() %>%
    ggplot(aes(x = date, y = steps)) +
        geom_col() +
        labs(x = 'Date', y = 'Number of steps')
```

### 2. Mean and median totals
```{r}
df %>%
    drop_na() %>%
    group_by(date) %>%
    summarise(
        steps = sum(steps)) %>%
    ungroup() %>%
    summarise(
        mean_steps = mean(steps),
        median_steps = median(steps))
```

## Average daily activity pattern
### 1. Timeseries plot
```{r}
df %>%
    group_by(interval) %>%
    summarise(
        mean_steps = mean(steps, na.rm=TRUE)) %>%
    ggplot(aes(x = interval, y = mean_steps)) +
        geom_line() + 
        labs(x = 'Time interval', y = 'Mean number of steps')
```

### 2. Which 5-minute interval contains the maximum mean number of steps?
```{r}
df %>%
    group_by(interval) %>%
    summarise(
        mean_steps = mean(steps, na.rm=TRUE)) %>%
        top_n(n=1)
```

## Imputing missing values
### 1. Calculate the total number of missing values
```{r}
sum(is.na(df))
```

### 2/3. Fill in missing values
Strategy for filling in missing values is to use the mean value for that
five-minute interval.
```{r}
# create reference data
interval_mean_steps <- df %>%
    group_by(interval) %>%
    summarise(
        mean_steps = mean(steps, na.rm=TRUE))

# loop over df to find rows where is.na(steps)==TRUE. 
# for those rows, get the interval value, but that into
# the reference data to find the row containing the relevant
# value of mean_steps for that interval, and insert that into
# the imputed df.

imputed_df <- df

for (row in 1:nrow(imputed_df)) {
    if (is.na(imputed_df[row, 'steps'])) {
        interval <- imputed_df[row, 'interval']
        new_row <- which(interval_mean_steps$interval==interval)
        mean_steps <- interval_mean_steps[new_row, 'mean_steps']
        imputed_df[row, 'steps'] <- mean_steps
    }
}
```

### 4. Histogram of total number of steps taken each day with imputed data, with mean and median numbers per day reported
```{r}
imputed_df %>%
    ggplot(aes(x = date, y = steps)) +
        geom_col() +
        labs(x = 'Date', y = 'Number of steps')

imputed_df %>%
    group_by(date) %>%
    summarise(
        steps = sum(steps)) %>%
    ungroup() %>%
    summarise(
        mean_steps = mean(steps),
        median_steps = median(steps))
```
Conclusion: This imputation approach made virtually no difference to the mean
and median steps per day (which is intuitive, given that the imputation just
used the mean values for a given interval). The main difference is that the
histogram appears more 'filled in', i.e., there are fewer days with all or nearly all missing
values (and hence low step counts).

## Are there differences in activity patterns between weekdays and weekends?
### 1. Mutate the data
```{r}
daytype_df <- imputed_df %>%
    mutate(
        daytype = case_when(
            weekdays(date) %in% unique(weekdays(imputed_df$date))[1:5] ~ 'Weekday',
            TRUE ~ 'Weekend')) 
```

### 2. Facet plot
```{r}
daytype_df %>%
    group_by(daytype, interval) %>%
    summarise(
        mean_steps = mean(steps, na.rm=TRUE)) %>%
    ggplot(aes(x = interval, y = mean_steps)) +
        geom_line() +
        labs(x = 'Time interval', y = 'Mean number of steps') +
        facet_wrap(~ daytype)
```