# Caching computations

- Store results of code in a database (i.e., cache them)

## cacher package for R
- evaluates code written in files and stores intermediate results in a key-value
  database
- R expressions are given SHA-1 hash values so that changes can be tracked and
  code reevaluated if necessary
- 'cacher packages' can be built for distribution
- others can 'clone' an analysis and evaluate subsets of code or inspect data
  objects

## Using cacher as an author
1. parse the R source file; create the necessary cache directories and
   subdirectories
2. cycle through each expression in the source file:
    - if an extression has never been evaluated, evaluate it and store any
      resulting R objects in the cache database
    - if a cached result exists, lazy-load the results from the cache database
      and move to the next expression
    - if an expression does not create any R objects (i.e., there is nothing to
      cache), add the expression to the list of expressions where evaluation
      needs to be forced
    - write out metadata for this expression to the metadata file

- the ```cachepackage``` function creates a ```cacher``` package storing
    - source file
    - cached data objects
    - metadata
- package file is zipped and can be distributed
- readers can unzip the file and immediately investigate its contents via
  ```cacher``` package

## Using cacher as reader

```{r}
library(cacher)
clonecache(id = "...")
showfiles()
sourcefile('top20.R') # e.g.
```

## Cloning an analysis
- local directories created
- source code files and metadata are downloaded
- data objects are not downloaded by default
- references to data objects and loaded and corresponding data can be lazy-loaded on demand

### tracing code backwards
```{r}
objectcode(data)
```
## Running code
- the runcode function executes code in the source file
- by deafult, expressions that result in an object being created are not run and the results objects are lazy loaded into the workspace
- expressions not results in objects are evaluated

## Checking code and objects
- the checkcode function evaluates all expressions from scratch (no lazy loading)
- results of evaluation are checked against stored results to see if the results are the same as what the author calculated (setting rng seeds is critical)
- the integrity of data objects can be verified with the ```checkobjects``` function to check for possible corruption of data (e.g., in transit)

## Inspecting data objects
```{r}
loadcache()
ls()
```

## Cacher summary
- the ```cacher``` package can be used by authors to create cache packages from data analyses for distribution
- readers can use the cacher package to inspect others' data analyses by examining cached computations
- cacher is mindful of readers' resources and efficiently loads only those data objects that are needed